\chapter{Conclusions and Future Work}
\label{chap:conclusions}

\section{Summary}
\label{sec:concl-summary}

AI shopping assistants have moved from research to production at
hyperscale, and the published evidence on the largest deployment
documents systematic failures in recommendation accuracy, factual
fidelity, and intrusiveness. The motivating question of this
thesis was whether those failures are baked into the AI-shopping
problem or whether they come from specific integration choices a
different system could have made differently. The work argues
for the second reading and builds a prototype that demonstrates
it.

The prototype combines a two-stage recommendation pipeline
(classical candidate generation followed by LLM rerank and
explanation), a reactive-first advisor surface driven by
lightweight behavior signals, a review-summarization module
grounded in real review text with explicit conflict surfacing,
an AI-assisted comparison flow with no winner verdict, and an
output-validation layer that strips fabricable claims from free
text and checks every cited identifier against the live catalog.
A custom evaluation harness sits alongside the prototype, scoring
configurations on grounded quality, latency, and cost across a
nine-configuration canonical sweep.

The measured evidence supports the central claim on two of three
axes. Every reported configuration completes well under the
consumer-tolerance latency bound (p95 below 8.1~s for the slowest
configuration; below 5~s for the seven non-worst configurations)
and at well under one cent per query, with grounded output
across the sweep (\texttt{groundedness} mean $\ge 0.89$ for every
configuration). The harness adds a finding the claim did not
anticipate: there is a real design space within those bounds
where the right choice is the cheaper, less computationally
aggressive configuration rather than the one with the highest
raw quality. Higher reasoning effort and looser step budgets are
uniformly worse on composite, which disproves the implicit
``more reasoning is always better'' assumption that production
code tends to bake in.

\section{Practical Implications}
\label{sec:concl-implications}

Five recommendations follow from the design and the measured
results.

\begin{description}
  \item[Never let the LLM generate factual product data.]
    Hydration-over-generation reduces the price-hallucination
    surface to zero by construction. Cited identifiers are
    verified against the live catalog; price and specification
    claims are stripped from free text before render.
  \item[Surface review conflicts explicitly with counts.] A
    summarized opinion that hides the dissenting minority is the
    wrong opinion when the minority is large enough to matter;
    the explicit-count format makes that visible.
  \item[Keep AI advice user-initiated.] Behavior signals prepare
    context but do not auto-fire the LLM. The reactive surface
    respects user agency and avoids the intrusiveness pattern
    documented for comparable production systems.
  \item[Route by query type.] The harness shows a real
    cost-quality tradeoff between Flash-Lite and
    \texttt{no-fewshot} \texttt{gpt-oss-120b}, and a uniform
    loss from higher reasoning effort. Configuration choices
    should be backed by measured evidence, not intuition.
  \item[Measure every configuration choice.] The few-shot
    contamination finding is the canonical example: the
    production UI hid the failure mode, and only a harness
    running against a broader query distribution surfaced it.
\end{description}

\section{Limitations}
\label{sec:concl-limitations}

The biggest scope limits are the absence of a human-subject
evaluation, the absence of an offline-IR comparison for the
classical recommendation algorithms, the seeded-catalog scale
(about 200 products, 4\,836 AI-generated reviews), the
single-deployment replication horizon, and the OpenRouter
\texttt{insufficient\_credits} caveat that biases the row
coverage of long-output configurations in the canonical sweep.
The thesis does not claim to address any of these; it claims
that the engineering work it does report holds up under
measurement within those limits.

\section{Future Work}
\label{sec:concl-future}

A within-subjects user study (perceived usefulness, trust,
intrusiveness, purchase confidence) is the single most valuable
extension: the engineering-quality metrics in
Chapter~\ref{chap:system} do not predict UX outcomes, and human
evaluation is what would close that gap. Adding offline
recommender-quality metrics (Hit~Rate@5, NDCG@10, catalog
coverage, intra-list diversity, popularity-bias Gini) against a
held-out temporal split of the seeded order data is a low-risk
extension that closes the second methodological gap, since it
shares the same data pipeline. Extending the harness to
multi-category catalogs and non-electronics domains would test
whether the theme-fidelity scorer and the comparison flow
generalize beyond the curated dataset. Richer behavior signals
(mouse trajectory, eye tracking, repeated-visit patterns) are
cheap to capture and worth evaluating once human-subject data is
available. Production hardening at 10\,000+ products and traffic
load is documented in \texttt{docs/scale-considerations.md};
evolving the eval harness into a CI regression gate with pareto
comparison against a baseline would let the design discipline
this thesis argues for survive contact with a real engineering
workflow.

\section{Closing Remarks}
\label{sec:concl-closing}

The thesis argues, end to end, that trustworthy AI shopping
assistance is a problem of integration discipline, not of model
capability. The failure modes documented for the largest
production system are addressable through structural decisions a
different system can make differently, and a working prototype
plus an evaluation harness are enough to show that those
decisions hold up under measurement.
