\chapter{Background and Related Work}
\label{chap:background}

\section{Classical Recommender Systems}
\label{sec:bg-recsys}

Three families dominate e-commerce recommendation: collaborative
filtering (with co-occurrence as its lightweight specialization),
content-based filtering, and hybrids of the two. Time-decayed
popularity sits alongside them as the workhorse for trending
slots. Each family is covered in detail
elsewhere~\cite{ComprehensiveRecSys2024, LLM4Rec2025}; the formal
definitions used in this thesis are in
Chapter~\ref{chap:foundations}.

\subsection{Co-Occurrence and Collaborative Filtering}
\label{subsec:bg-cooccurrence}

Item-based collaborative filtering uses the simple regularity
that users who buy product $A$ often also buy product $B$. The
core artifact is a co-purchase matrix that counts the buyers
shared by each pair of items. A similarity score is derived from
it, and Jaccard is the most common lightweight choice because it
normalizes by union and is robust to differences in product
popularity~\cite{ComprehensiveRecSys2024}. Item-based variants
win at e-commerce scale because product catalogs have more buyers
per item than items per buyer, which makes the item similarity
matrix denser and more stable. The main cost is precomputation:
production deployments compute the matrix on a schedule, not per
request.

The strengths are interpretability, no model training, and
tolerance to sparse user history. The weaknesses are blindness to
product attributes (two unrelated products that share a buyer
base still link) and the cold-start problem for new products with
no purchase history.

\subsection{Content-Based Filtering}
\label{subsec:bg-content}

Content-based filtering scores similarity using product
attributes instead of purchase signals. A typical setup defines a
feature vector over category, brand, tags, and price proximity,
then computes a weighted distance between the active product and
every other product in the catalog. The main upside is the
zero-purchase cold start: a brand-new product can be recommended
on day one because its attributes are known immediately. The main
downside is the filter-bubble effect: recommendations reinforce
the active item's profile.

\subsection{Hybrid Recommenders and Trending}
\label{subsec:bg-hybrid}

Burke's taxonomy splits hybrid recommenders into seven patterns,
of which weighted, switching, and mixed are the ones most often
deployed~\cite{Burke2002Hybrid}. A weighted hybrid combines
normalized scores from multiple algorithms into a single ranking.
A switching hybrid picks which algorithm to run per request based
on context (cold versus warm user, sparse versus rich product). A
mixed hybrid shows results from multiple algorithms side by side
without combining their scores. The prototype uses a switching
hybrid for the personalized ``recommended for you'' slot. Where
scores are combined, the prototype uses Reciprocal Rank Fusion,
which sidesteps the problem of normalizing scores across
different algorithms by combining ranks instead~\cite{LLM4Rec2025}.

Naive popularity over an order log favors products with long-tail
demand and penalizes products that are gaining fast. Exponential
time decay corrects for this by weighting recent activity more
heavily, with a half-life chosen per category: short for fashion
and consumer trends, long for durable
goods~\cite{SessionRecSys2021}. The formal definition is in
Section~\ref{sec:found-similarity}.

\section{Explainable Recommendations}
\label{sec:bg-explainable}

A recommendation with a reason is a different thing from one
without. Classical systems offer pattern-level explanations
(``users who bought X also bought Y'') that are interpretable but
generic. Recent work uses LLMs to generate sentence-level
rationales conditioned on user
context~\cite{LLMEnhancedRecSys2024}. Knowledge-graph-grounded
explanations sit between the two: they draw facts from a curated
graph instead of generating freely, and they reduce hallucinated
attributes in the rationales they
produce~\cite{ExplainableRecKG2024}. Having an explanation
improves trust calibration: users accept correct recommendations
more readily and reject incorrect ones more readily. The thesis
treats explanation as a first-class output of the LLM stage.

\section{Large Language Models in Recommender Systems}
\label{sec:bg-llm-recsys}

Surveys of LLM-augmented recommendation identify three roles for
an LLM in the pipeline: knowledge enhancement (the LLM acts as a
knowledge source for product attributes and user intent),
interaction (the LLM is the conversational interface to a
recommender backbone), and model enhancement (LLM-derived
features feed into a downstream ranker as extra
input)~\cite{LLMEnhancedRecSys2024}. The roles differ in cost and
integration depth. Interaction is the cheapest to deploy because
it sits at the surface and needs no retraining. Model enhancement
is the most invasive because it changes the offline training
pipeline. This thesis uses roles 1 and 2: the LLM contributes
knowledge (review themes, comparison narratives, product
explanations) and interaction (the conversational advisor). It
does not retrain a ranker. The data volume of a bachelor-thesis
prototype would not support it, and the architectural lessons of
the thesis are at the integration layer, not the modeling layer.

The standard pattern for combining classical recommenders with
LLMs is a two-stage architecture: a classical candidate generator
returns a small set of products under a tight latency budget, and
an LLM reranks, explains, or summarizes that set in a second
stage~\cite{HSTU2024, ShopifyHSTU}. The pressure is cost: LLM
inference scales with context size, so feeding five candidates is
fine at consumer scale, while feeding five thousand catalog items
is not. The pattern also separates concerns cleanly. The
classical stage owns relevance and recall over the full catalog;
the LLM stage owns explanation, conflict surfacing, and tradeoff
articulation. The prototype follows this split strictly: the
classical stage runs on the full Convex catalog, and the LLM
stage only ever sees a small candidate set, never the full
catalog.

\section{Review Summarization for Purchase Validation}
\label{sec:bg-review-summarization}

Reviews drive a large share of purchase decisions, yet most
product detail pages show them raw. AI-generated review summaries
compress the work of reading them into a short structured
artifact: themes, frequently mentioned features, and (less often)
explicit conflict callouts.

Naive summarization fails in two specific ways. First, it
averages sentiment, hiding the minority opinions that are often
the most useful for purchase decisions (a 38-versus-9 split
between buyers who love a product and buyers who report failure
after six months is more useful than a 4.2-star average). Second,
when prompted to produce a tidy summary, it tends to fabricate
themes that are not in the source
corpus~\cite{RufusConfidentlyWrong2024}. The second failure has
drawn the most criticism of comparable production review-summary
systems. The thesis treats theme fabrication as a hard
constraint: summaries are generated against the actual review
corpus, themes are validated against the source text before
persistence, and conflicting opinions are surfaced explicitly
with counts.

\section{AI-Assisted Comparison as Decision Support}
\label{sec:bg-comparison}

Comparison is the highest-cognitive-load task in a typical
purchase journey. Users compare across mixed attributes (price,
battery life, build material, warranty length, shipping speed)
where the weight of each attribute depends on personal
preferences. LLMs are well suited to laying out these tradeoffs
in natural language. Industry agrees: Shopify's analysis of
agentic shopping flows names comparison and tradeoff articulation
as two of the interactions where LLMs add the most user-perceived
value~\cite{ShopifyAgentic}. Consumer research finds 56\% of
shoppers explicitly want comparison help from AI assistants and
47\% want review summarization~\cite{DigidayAIShopping}.

The main design failure in current AI comparison features is the
``best product'' verdict: the system picks a winner, hides the
tradeoffs, and takes the choice away from the user. UX guidance
for agentic AI explicitly warns against this pattern. It
recommends a tradeoff-first layout where the system surfaces the
structured differences and lets the user weight
them~\cite{SmashingAgenticUX}. The thesis follows the
tradeoff-first pattern: comparisons present structured
differences, never a verdict.

\section{Trust and Human-AI Interaction}
\label{sec:bg-trust}

\subsection{Timing and Consumer Resistance}
\label{subsec:bg-timing}

A CHI 2024 study measured how users react to AI suggestions
delivered at different timings and found that suggestions offered
after a natural pause were rated as significantly more helpful
than suggestions injected during active
deliberation~\cite{CHI2024Timing}. A related body of work in
human-AI personalization makes a stronger claim: reactive
personalization preserves the user's perceived autonomy, and that
perceived autonomy is one of the strongest predictors of trust in
AI systems~\cite{ProactiveReactivePersonalization}. Proactive
interventions can be welcome, but only at high-confidence moments
where the system has strong evidence the user is genuinely stuck
or at decision risk.

The market data on consumer trust in AI shopping assistants is
low. 55\% of consumers say they do not trust AI chatbot product
recommendations~\cite{ConsumerDistrust2025}. 42\% see e-commerce
AI assistants as upsell tools, not advisors~\cite{AIUpsell2025}.
71\% of retailers have deployed conversational AI in some form,
but only 16\% of consumers use these assistants
regularly~\cite{ConsumerResistanceAI}. The gap between deployment
and usage points to structural resistance. A trustworthy AI
advisor is not a polish layer on top of a shopping system; the
trust gap is the product question.

\subsection{Behavioral Signals as Intent Indicators}
\label{subsec:bg-behavior-signals}

A long line of work in information retrieval establishes
lightweight client-side signals (cursor position, dwell time,
scroll depth, hover trajectory) as reliable proxies for user
attention~\cite{SIGIR2020Cursor}. The signals are cheap to
collect and need no server-side modeling at capture time. Dwell
time deserves a note. In information retrieval, dwell predicts
engagement well: the longer a user reads, the more engaged they
are. In e-commerce, dwell still predicts engagement, but it
predicts intent-to-buy less reliably; users sometimes dwell
because they are confused, comparing across browser tabs, or
distracted~\cite{DwellTimeEngagement2022}. The thesis treats
dwell as an attention proxy, not a buying-intent proxy, and
combines it with review-tab engagement and scroll depth in a
composite readiness score.

A privacy boundary applies. The prototype aggregates raw signals
client-side and persists only summary statistics to the backend.
Raw event streams are never stored or transmitted off the device.

\section{Case Study: Amazon Rufus}
\label{sec:bg-rufus}

Amazon Rufus is the largest deployed AI shopping assistant by
user count and the most thoroughly studied externally. Rufus is
the reference point for this section: its architecture documents
the practical shape of a production-scale AI shopping system, and
its documented failures inform several specific design decisions
in the thesis prototype.

\subsection{System Architecture and Capabilities}
\label{subsec:bg-rufus-arch}

Rufus combines three architectural patterns. A multi-model router
picks an inference model per query: simple lookups go to a fast
cheap model, while reasoning-heavy queries go to a slower, more
capable one. The output is two-stream: text is streamed as it
generates, and structured product references are hydrated
client-side from live catalog data, so the user sees the
explanation and the live product card as a single coherent
response~\cite{RufusTechnology, RufusScaling}. The system plugs
into Amazon's COSMO knowledge graph, an LLM-built commonsense
graph that enriches products with implicit attributes (a
children's bicycle is also ``a gift for a 7-year-old''),
reportedly producing a 60\% improvement on internal relevance
metrics~\cite{COSMO}. The integration footprint is broad: Rufus
appears on the product detail page, in search, and in the cart,
and serves over 300 million users with parallel speculative
decoding to keep latency within product
targets~\cite{RufusScaling}.

\subsection{Documented Accuracy and Trust Issues}
\label{subsec:bg-rufus-problems}

External evaluations of Rufus report four classes of failure,
stable across studies. First, recommendation accuracy: independent
measurement places Rufus at 32\% accuracy for picking the right
product for a query, far below the relevance figures Amazon
reports internally~\cite{RufusWrong2024}. Second, factual
hallucination: 28\% of Rufus responses that include a product
price quote the wrong price, an error type that is especially
damaging because the user can check it in
seconds~\cite{RufusWrong2024}. Third, self-serving bias: 83\% of
Rufus's recommendations favor Amazon-owned or Amazon-branded
products over third-party alternatives, a pattern that reinforces
the perception of the assistant as an upsell
tool~\cite{RufusWrong2024}. Fourth, review summary fabrication:
Rufus has been observed to generate review themes that do not
appear in the source review corpus, the failure that has drawn
the most public criticism of the
four~\cite{RufusConfidentlyWrong2024}.

The four failure modes are not independent. They share a single
root cause: the LLM is allowed to generate factual product data
(prices, attributes, themes) from its parametric memory instead
of being constrained to read from live catalog data and source
review text. This observation directly motivates the output
validation and hydration strategy in
Chapter~\ref{chap:foundations}.

\subsection{Feature Evolution and Lessons}
\label{subsec:bg-rufus-lessons}

Rufus's product timeline is informative beyond the architecture.
The system launched in early 2024 as a reactive assistant: the
user opened a chat surface and the system responded. The
proactive ``Help Me Decide'' surface, where the system offers to
compare or recommend without being asked, was added about ten
months later~\cite{RufusTechnology, RufusScaling}. The phasing
matches the CHI 2024 timing finding above: reactive came first
because it was easier to launch without breaking user
expectations, and proactive was added piece by piece once the
system had stabilized. Public reporting credits Rufus with
roughly \$12 billion of incremental sales in
2025~\cite{RufusRevenue2025}; the thesis treats this number as a
market-sizing data point, not proof of product-market fit, since
the trust failures above leave the user-experience case
contested.

\section{Positioning of This Thesis}
\label{sec:bg-positioning}

Prior work treats classical recommendation and LLM integration
as separate problems. The questions that decide whether a real
AI shopping advisor is trustworthy (when to speak, how to ground
its answers, how to surface conflicting reviews, how to catch
hallucinations) sit scattered across HCI papers, engineering
blogs, and post-mortems of shipped systems. This thesis pulls
them into one design, tests them on a working prototype, and
backs every model and prompt choice with evidence from a custom
evaluation harness. The contribution is not a new algorithm. It
is an integration architecture and an honest evaluation against
documented production failures, at a scale a bachelor thesis can
defend.
